\Title

### 2023-02-03 Grimaldi-etal-BiolCybernetics.tex
figure comments

\subsection{Detecting event-based motifs using spiking neurons with heterogeneous delays}% model as a Logistic Regression with a temporal convolution}

@@

@@

\begin{figure}%[t!]
    \includegraphics[width=0.490\linewidth]{figures/THC_1a_k.pdf}
    \includegraphics[width=0.490\linewidth]{figures/THC_1c.pdf}
    \includegraphics[width=0.490\linewidth]{figures/THC_1b.pdf}
    \includegraphics[width=0.490\linewidth]{figures/THC_1a.pdf}
	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 
	    {\bf (a)}~Given a generic raster plot defined by a set of spikes occurring on specific neural addresses and at specific times, one may consider that this information consists of the repeated occurrence of precise spiking motifs. 
	    {\bf (b)}~We show the motifs used in this example, each identified at the top by a different color. To each of the $10$ neural adresses and $71$ different possible delays is assigned an evidence of activation (red) or deactivation (blue). Note that each afferent may be connected with multiple weights at different delays.
	    {\bf (c)}~The activation in time of the different motifs is then used to define a generative model for drawing a raster plot on the multi-unit address space: The propagation of the afferent information through these delays generates the raster plot seen in (a). 
	    {\bf (d)}~The inversion of the generative model provides with a model which gives the predicted probability of occurrence of each motif at any time and which may be used to generate a spike as a Bernoulli trial. The detection model yields in this particular case with an exact identification of the occurrence of motifs. Knowing the results of this detection, one may for illustration purposes highlight them by different colors in the raster plots, showing that in this synthetic examples, all spikes can be annotated with each identified spiking motif. 

@@

\subsubsection{A generative model for raster plots}

@@

In neurobiological recordings or in the sensory signal obtained from an event-based camera, any generic raster plot consists of a stream of \emph{spikes} or \emph{events} (see Figure~\ref{fig:model}-(a)). This can be formalized as a list of neural addresses and timestamps tuples $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events. Events are typically ordered by their time of occurrence. Each event has a time of occurrence $\timev_\arank$  and an associated address $\presynaddr_\arank$ (which is typically in the form $(x_\arank, y_\arank, \polev_\arank)$ for event-based cameras). This defines an address space $\presynaddrspace$ which consists of the set of possible addresses. In a neurobiological recording, this can be the identified set of neurons. For event-based cameras, it is denoted by $[1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities coded in event-based cameras).

% This raster plot is generated by spiking neurons which are defined on the one hand by the equations governing the evolution of its membrane potential dynamics on its soma and on the other hand by the characterization of the synaptic contacts on its dendritic tree, or on the sensory processes (such as the pixel on the sensor of a camera, or the photoreceptors in the retina). A classical characterization consists in detailing the synaptic weights of each synaptic contact on the dendritic tree, the so-called weight matrix. A more detailed description of these synaptic contacts reveals the added importance of heterogeneous delays, i.e., the precise timing from one afferent neuron's firing to its arrival in the soma, and how this changes the network's dynamics~\citep{izhikevich_polychronization_2006}. In that description, we can parameterize each neuron by the set of tuples defining both the weight and the delay of each synaptic contact. As a consequence, a set of input presynaptic spikes $\event$ will be processed by the dendrites defined by this synaptic set and notably by the respective delays, which will multiplex in time all events.

Let's formalize a layer of spiking neurons with heterogeneous delays (HD-SNN). Each neuron $\postsynaddr \in \postsynaddrspace$  connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses. Each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. Scanning all neurons $\postsynaddr$, we thus define the full set of $\Nsyn$ synapses, 
as  $\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$, where each synapse is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$,  a weight $\synapticweight^\ranksyn$, and a delay $\synapticdelay_\ranksyn$. This defines the full connectivity of the HD-SNN model. Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]} $, or the emitting field of a presynaptic neuron $\synapse_\presynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]}$. As a consequence, an event stream which evokes neurons in the presynaptic address space is multiplexed by the synapses into a new event stream which is defined by the union of the sets generated by each emitting field from the presynaptic space: 
$ \cup_{\arank \in [1,\numevent]} \{ \{(\postsynaddr_\ranksyn, \synapticweight_\ranksyn, \timev_\arank + \synapticdelay_\ranksyn) \}_{ \ranksyn \in \synapse_{\presynaddr_\arank}} \}$. This new stream of events is by nature ordered in time as events reach the soma of post-synaptic neurons. In particular, when post-synaptic neurons are activated on their soma by this spatio-temporal motif, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner. 
% \note{ TODO: show why this formula implements a polychrony detector }

% \note{ éclaircir }

Taking the argument the other way around, one may from a generative model for generic raster plots. Indeed, any spike in the presynaptic address space is generated by sensory neurons (for instance photoreceptors in the retina, sensors in a CMOS chip) or by afferent spiking neurons. In the latter case, these are connected to the spiking cell by a set of weights and delays, whose structure is stable relatively to the coding timescale. When these connections are high and sparsely distributed, this firing will cause a specific temporal motif. Another example is given for the barn owl auditory system: As it hears the sound of a mouse, this sound will generate a specific spiking response in both ears, and specifically, the precise timing between the signal generated by the left relative to the right ear can be for instance used to determine the position of the prey~\citep{goodman_spike-timing-based_2010}. Overall, these examples show that raster plots may be considered as a mixture of the effects of different elementary causes, and that each event triggers a specific spatio-temporal spiking motif. 
%
\subsubsection{Detecting spiking motifs}
%: Detection model

From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers, it is useful to transform this event-based representation into a dense representation. Indeed, we may transform any event-based input as the boolean matrix $A \in \{0, 1 \}^{N\times T}$, where $N$ is the number of neurons in $\presynaddrspace$ and $T$ is the number of time bins. In this simplified model, we will consider that heterogeneous delays are limited in range such that the synaptic set can be represented by the dense matrix $\kernel^\postsynaddr$ giving for each neuron $\postsynaddr$ the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in [1,\Nsyn^\postsynaddr]}, \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) = \synapticweight_\ranksyn^\postsynaddr$. 
The probability of firing of a neuron $a$ at a given time $t$ can be understood as a Bernoulli trial whose (only) parameter is a bias $p(t, a) \in [0, 1]$. Assuming that the presence of spiking motifs conditions the probability on all efferents, the logit (inverse of the sigmoid) of this probability bias can be written as the sum of the logit of each of these factors, whose values are defined by the corresponding weights. Spiking motifs may be activated independently and at random times, such that  we write this activity as $B(b, t)=1$ if $b$ is activated at $t$ and else $B(b, t)=0$. We can thus write the probability bias as the accumulated evidence given these factors as 
\begin{equation*}
p(t, a) = \sigma\big(W_0 + \sum_{b, t} B(b, t) \cdot W_b(a, t-d) \big)  
\end{equation*}
where $\sigma$ is the sigmoid function. We will further assume that the weights are balanced (their mean is zero) and that $W_0$ is a bias such that $p_0=\sigma(W_0)$ is the average background firing rate. Conveniently, one can write this summation as a one-dimensional temporal convolution operator such that we may simply write
\begin{equation*}
p = \sigma(W_0 + B \ast W )
\end{equation*}
where  $p\in [ 0, 1]^{N\times T}$ and $B\in \{0, 1\}^{M\times T}$ is the raster plot corresponding to the temporal activation of the PGs. Finally, we obtain the raster plot $A\in \{0, 1\}^{N\times T}$ by drawing spikes using independent Bernoulli trials $A \sim \mathcal{B}(p)$. Note that, depending on the shape of the kernels, the generative model can model a discretized Poisson process, generate rhythmic activity or more generally propagating waves. This formulation thus defines a simple generative model for raster plots as a combination of independent PGs. 

@@

% 
This generative model defined above allows to determine this inference model for guessing sources $B$ when observing a raster plot $A$. This assumes that we know the spiking motifs as defined by the $W_b$ matrices. The underlying metric is the binary cross-entropy, as used in the logistic regression model. In particular, if we consider kernels with similar decreasing exponential time profile, we can prove that this is similar to the method of~\citet{berens_fast_2012}. In our specific case, the difference is that the regression is performed in both dendritic and delay space by extending the summation using a temporal convolution operator. Using this forward model, it is possible to estimate the logit (inverse of a sigmoid) $\hat{B}(b, t)$ for the presence of a PG of address $b$ and at time $t$ by using the transpose convolution operator. Equivalently, this consists in using the emitting field $\synapse_\presynaddr$ of presynaptic neurons in place of the receptive field $\synapse^\postsynaddr$ of postsynaptic neurons. It thus comes that when observing $A$, then one may infer $\hat{B} = A \ast W^T$ and select the most activated items. This assumption hold as long as the kernels are uncorrelated, a condition which is met here numerically by chosing a relatively sparse set of synapses (approximately $1\%$ of active synapses).

%\subsubsection{Implementation using spiking neurons with heterogeneous delays}
%
% The activation function of our spiking neural is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
%It transforms this list of weights into a probability with the following formula:
%$
%Pr(k=\postsynaddr \; \vert \; \timev) =
%\frac 1 Z
%{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
%$ 
%where $\mathcal{C}^\postsynaddr(\timev) = \sum
%\activeweights^\postsynaddr(t)
%$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
%In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%%
%
% \note{say that it is effortless in biology but difficult on conventional computers}
% In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the heterogeneous delays model. 
%
\subsubsection{Performance on synthetic data}
%
\begin{figure}%[t!]
    \centering
    \includegraphics[width=0.490\linewidth]{figures/THC_N_PGs.pdf}
    \includegraphics[width=0.490\linewidth]{figures/THC_N_PG_time.pdf}
	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 
	    {\bf (a)}~Accuracy of detection as a function of the number of kernels.
	    {\bf (b)}~Accuracy of detection as a function of the temporal depth $D$ of kernels among $M=144$ kernels.
	    }
    \label{fig:model_results}
\end{figure}

To quantify the efficiency of this operation, we generated $M=144$ synthetic spiking motifs as random independent kernels over $128$ presynaptic inputs and $D=71$ possible delays. We drew random independent instances of $B$ with a length of $T=1000$ time steps and with on average $1.0$ occurrences in each draw. This allowed us to generate raster plots which we use to infer $\hat{B}$. We compute the accuracy as the rate of true positive detections (both for inferring the address and its exact timing) and observe on average $\approx 98\%$ correct detections. We further extended this result by showing how the accuracy would evolve as a function of the number of simultaneous PGs, while keeping the same frequency of occurrence. We show in Figure~\ref{fig:model_results}~(a) that the accuracy of finding the right PG is still above $80\%$ accuracy with more than $1364$ overlapping PGs. Moreover, we show in Figure~\ref{fig:model_results}~(b) that (with $M=144$ PGs fixed) the accuracy increases notably as the temporal depth $D$ of the PG kernel increased, demonstrating quantitatively the computational advantage of using heterogeneous delays. These results were obtained while assuming that we know $W$. However, this is in general not the case, for instance when observing the raster plot of a population of neurons. In the following, we will define a generic visual task and determine a learning algorithm to solve that challenging task. %Inspired by the k-means algorithm, it is possible to devise a self-supervised learning algorithm. Our preliminary results show that it is possible to retrieve PGs embedded in the data, yet that further analysis is necessary to improve the convergence of the algorithm. In particular, it seems promising to use a sparseness constraint in the inference mechanism such as to remove spurious correlations in the inference.

>>>>

\subsection{Task definition: fast motion detection}

@@

Let us now define a procedure of animating of natural scene by virtual eye-movements  similar to that which was used in neurobiological~\citep{vinje_sparse_2000, baudot_animation_2013} and computational neuroscience~\citep{kremkow_push-pull_2016} studies. First, let's define a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the field of vision. In animals with a fovea, this is particularly useful as it allows to move the area with the highest density of photoreceptors in the environment, for example at a point of interest. A specific mechanism to do so are called saccades which are rapide eye movements of the eyeball that reposition the center of vision. In humans, these are very frequent (on average $2$ each second~\citep{dandekar_neural_2012}). They are produced very rapidly (about $80~\ms$) and display a whole range of speeds. At a more microscopic scale, the human gaze moves with an incessant drift similar to a Brownian-like trajectory~\citep{poletti_head-eye_2015}. To maintain the full generality of the task, we will define eye movements using a form of random walk~\citep{engbert_integrated_2011}. This approach first defines a finite set of possible 2D motions in polar coordinates. Based on the distribution of biological motions, we simplified it by selecting a set of eye movements as the Cartesian product of % N-V-phi
$8$ linearly spaced motion directions and %N-Y-velocity=
$7$ different velocities (\seeFig{motion_task}-(a)-inset). Note that the velocities are sampled on a geometric scale between $1/4$ and $4$. Next, we define one trajectory of gaze as segments whose duration is drawn from a Poisson distribution with an average block length of $20~\ms$, similarly to a Lévy flight~\citep[p. 289]{mandelbrot_fractal_1982}. Finally, the trajectory is integrated by assuming first that velocities are uniformly and independently sampled from the set of different motion-sets and second that motion is uniform during a time segment. The resulting instances yield trajectories qualitatively similar to those observed for human eye movements (\seeFig{motion_task}-(a)). 
%: fig:motion_task
\begin{figure}%[h!]%[ht!]
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    {\bf - Motion detection task.} {\bf (Left)}~We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view such as to reproduce the effect of displacing the eye. To mimic the effect of a saccadic-like eye movement, the axis of view moves following a step-like random walk, and we show here an example path of the trajectory as a green line ($200$ time steps). {\bf (Right)}~Snapshot of the synthesized event stream at different time steps. The dynamics of the sub-image as a function of time generates a naturalistic movie which may be transformed into an event-based representation. Mimicking the retina, this representation codes for proportional and pixel-wise increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. On the bottom left corner of the images, the translation vector are plotted in red among the different possible classes of movements. Note the change of direction between the second and third frames.}
    \label{fig:motion_task}
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (\seeFig{motion_task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration $N_T = 250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a motion of one pixel per frame (i.e., per time bin).

% \note{ TODO: formula / say it is driven by image gradients / give scaling rules }
To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(c)). The goal here is to infer the correct motion solely by observing these events. 
This sensory signal representing the output of this event-based camera forms a discrete stream of events, $\event$ as defined above.   
Each event has an associated address, which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$. This defines a presynaptic address space $\presynaddrspace = [1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities  ($\Npol=2$ for ON and OFF polarities). 
% \note{ the task is full-field rigid translations, the model resolves it locally (analogy to the input from V1 to MT) }

@@

%\subsection{Detecting event-based motifs using spiking neurons with heterogeneous delays}% model as a Logistic Regression with a temporal convolution}
%%%%-----------------------------------------------------------------
%\begin{figure}%[t!]
%    \centering
%    \includegraphics[width=0.490\linewidth]{figures/THC_1a_k.pdf}
%    \includegraphics[width=0.490\linewidth]{figures/THC_1c.pdf}
%    \includegraphics[width=0.490\linewidth]{figures/THC_1b.pdf}
%    \includegraphics[width=0.490\linewidth]{figures/THC_1a.pdf}
%	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 
%	    {\bf (a)}~Given a generic raster plot defined by a set of spikes occurring on specific neural addresses and at specific times, one may consider that this information consists of the repeated occurrence of precise spiking motifs. 
%	    {\bf (b)}~We show the motifs used in this example, each identified at the top by a different color. To each of the $10$ neural adresses and $71$ different possible delays is assigned an evidence of activation (red) or deactivation (blue). Note that each afferent may be connected with multiple weights at different delays.
%	    {\bf (c)}~The activation in time of the different motifs is then used to define a generative model for drawing a raster plot on the multi-unit address space: The propagation of the afferent information through these delays generates the raster plot seen in (a). 
%	    {\bf (d)}~The inversion of the generative model provides with a model which gives the predicted probability of occurrence of each motif at any time and which may be used to generate a spike as a Bernoulli trial. The detection model yields in this particular case with an exact identification of the occurrence of motifs. Knowing the results of this detection, one may for illustration purposes highlight them by different colors in the raster plots, showing that in this synthetic examples, all spikes can be annotated with each identified spiking motif. 
%	    }
%    \label{fig:model}
%\end{figure}
%\subsubsection{A generative model for raster plots}
%In neurobiological recordings or in the sensory signal obtained from an event-based camera, any generic raster plot consists of a stream of \emph{spikes} or \emph{events} (see Figure~\ref{fig:model}-(a)). This can be formalized as a list of neural addresses and timestamps tuples $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events. Events are typically ordered by their time of occurrence. Each event has a time of occurrence $\timev_\arank$  and an associated address $\presynaddr_\arank$ (which is typically in the form $(x_\arank, y_\arank, \polev_\arank)$ for event-based cameras). This defines an address space $\presynaddrspace$ which consists of the set of possible addresses. In a neurobiological recording, this can be the identified set of neurons. For event-based cameras, it is denoted by $[1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities coded in event-based cameras). 

@@

%% This raster plot is generated by spiking neurons which are defined on the one hand by the equations governing the evolution of its membrane potential dynamics on its soma and on the other hand by the characterization of the synaptic contacts on its dendritic tree, or on the sensory processes (such as the pixel on the sensor of a camera, or the photoreceptors in the retina). A classical characterization consists in detailing the synaptic weights of each synaptic contact on the dendritic tree, the so-called weight matrix. A more detailed description of these synaptic contacts reveals the added importance of heterogeneous delays, i.e., the precise timing from one afferent neuron's firing to its arrival in the soma, and how this changes the network's dynamics~\citep{izhikevich_polychronization_2006}. In that description, we can parameterize each neuron by the set of tuples defining both the weight and the delay of each synaptic contact. As a consequence, a set of input presynaptic spikes $\event$ will be processed by the dendrites defined by this synaptic set and notably by the respective delays, which will multiplex in time all events. 

@@

%Let's formalize a layer of spiking neurons with heterogeneous delays (HD-SNN). Each neuron $\postsynaddr \in \postsynaddrspace$  connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses. Each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. Scanning all neurons $\postsynaddr$, we thus define the full set of $\Nsyn$ synapses, 
%as  $\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$, where each synapse is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$,  a weight $\synapticweight^\ranksyn$, and a delay $\synapticdelay_\ranksyn$. This defines the full connectivity of the HD-SNN model. Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]} $, or the emitting field of a presynaptic neuron $\synapse_\presynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]}$. As a consequence, an event stream which evokes neurons in the presynaptic address space is multiplexed by the synapses into a new event stream which is defined by the union of the sets generated by each emitting field from the presynaptic space: 
%$ \cup_{\arank \in [1,\numevent]} \{ \{(\postsynaddr_\ranksyn, \synapticweight_\ranksyn, \timev_\arank + \synapticdelay_\ranksyn) \}_{ \ranksyn \in \synapse_{\presynaddr_\arank}} \}$. This new stream of events is by nature ordered in time as events reach the soma of post-synaptic neurons. In particular, when post-synaptic neurons are activated on their soma by this spatio-temporal motif, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner. 
%% \note{ TODO: show why this formula implements a polychrony detector }
%% \note{ éclaircir }
%Taking the argument the other way around, one may from a generative model for generic raster plots. Indeed, any spike in the presynaptic address space is generated by sensory neurons (for instance photoreceptors in the retina, sensors in a CMOS chip) or by afferent spiking neurons. In the latter case, these are connected to the spiking cell by a set of weights and delays, whose structure is stable relatively to the coding timescale. When these connections are high and sparsely distributed, this firing will cause a specific temporal motif. Another example is given for the barn owl auditory system: As it hears the sound of a mouse, this sound will generate a specific spiking response in both ears, and specifically, the precise timing between the signal generated by the left relative to the right ear can be for instance used to determine the position of the prey~\citep{goodman_spike-timing-based_2010}. Overall, these examples show that raster plots may be considered as a mixture of the effects of different elementary causes, and that each event triggers a specific spatio-temporal spiking motif. 
%\subsubsection{Detecting spiking motifs}
%%: Detection model
%From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers, it is useful to transform this event-based representation into a dense representation. Indeed, we may transform any event-based input as the boolean matrix $A \in \{0, 1 \}^{N\times T}$, where $N$ is the number of neurons in $\presynaddrspace$ and $T$ is the number of time bins. In this simplified model, we will consider that heterogeneous delays are limited in range such that the synaptic set can be represented by the dense matrix $\kernel^\postsynaddr$ giving for each neuron $\postsynaddr$ the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in [1,\Nsyn^\postsynaddr]}, \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) = \synapticweight_\ranksyn^\postsynaddr$. 
%The probability of firing of a neuron $a$ at a given time $t$ can be understood as a Bernoulli trial whose (only) parameter is a bias $p(t, a) \in [0, 1]$. Assuming that the presence of spiking motifs conditions the probability on all efferents, the logit (inverse of the sigmoid) of this probability bias can be written as the sum of the logit of each of these factors, whose values are defined by the corresponding weights. Spiking motifs may be activated independently and at random times, such that  we write this activity as $B(b, t)=1$ if $b$ is activated at $t$ and else $B(b, t)=0$. We can thus write the probability bias as the accumulated evidence given these factors as 
%p(t, a) = \sigma\big(W_0 + \sum_{b, t} B(b, t) \cdot W_b(a, t-d) \big)  
%where $\sigma$ is the sigmoid function. We will further assume that the weights are balanced (their mean is zero) and that $W_0$ is a bias such that $p_0=\sigma(W_0)$ is the average background firing rate. Conveniently, one can write this summation as a one-dimensional temporal convolution operator such that we may simply write
%p = \sigma(W_0 + B \ast W )
%where  $p\in [ 0, 1]^{N\times T}$ and $B\in \{0, 1\}^{M\times T}$ is the raster plot corresponding to the temporal activation of the PGs. Finally, we obtain the raster plot $A\in \{0, 1\}^{N\times T}$ by drawing spikes using independent Bernoulli trials $A \sim \mathcal{B}(p)$. Note that, depending on the shape of the kernels, the generative model can model a discretized Poisson process, generate rhythmic activity or more generally propagating waves. This formulation thus defines a simple generative model for raster plots as a combination of independent PGs. 

@@

% 
%This generative model defined above allows to determine this inference model for guessing sources $B$ when observing a raster plot $A$. This assumes that we know the spiking motifs as defined by the $W_b$ matrices. The underlying metric is the binary cross-entropy, as used in the logistic regression model. In particular, if we consider kernels with similar decreasing exponential time profile, we can prove that this is similar to the method of~\citet{berens_fast_2012}. In our specific case, the difference is that the regression is performed in both dendritic and delay space by extending the summation using a temporal convolution operator. Using this forward model, it is possible to estimate the logit (inverse of a sigmoid) $\hat{B}(b, t)$ for the presence of a PG of address $b$ and at time $t$ by using the transpose convolution operator. Equivalently, this consists in using the emitting field $\synapse_\presynaddr$ of presynaptic neurons in place of the receptive field $\synapse^\postsynaddr$ of postsynaptic neurons. It thus comes that when observing $A$, then one may infer $\hat{B} = A \ast W^T$ and select the most activated items. This assumption hold as long as the kernels are uncorrelated, a condition which is met here numerically by chosing a relatively sparse set of synapses (approximately $1\%$ of active synapses).

%\subsubsection{Implementation using spiking neurons with heterogeneous delays}
% The activation function of our spiking neural is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
%It transforms this list of weights into a probability with the following formula:
%$
%Pr(k=\postsynaddr \; \vert \; \timev) =
%\frac 1 Z
%{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
%$ 
%where $\mathcal{C}^\postsynaddr(\timev) = \sum
%\activeweights^\postsynaddr(t)
%$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
%In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
% \note{say that it is effortless in biology but difficult on conventional computers}
% In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the heterogeneous delays model. 
%\subsubsection{Performance on synthetic data}
%\begin{figure}%[t!]
%    \centering
%    \includegraphics[width=0.490\linewidth]{figures/THC_N_PGs.pdf}
%    \includegraphics[width=0.490\linewidth]{figures/THC_N_PG_time.pdf}
%	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 
%	    {\bf (a)}~Accuracy of detection as a function of the number of kernels.
%	    {\bf (b)}~Accuracy of detection as a function of the temporal depth $D$ of kernels among $M=144$ kernels.
%	    }
%    \label{fig:model_results}
%\end{figure}
%To quantify the efficiency of this operation, we generated $M=144$ synthetic spiking motifs as random independent kernels over $128$ presynaptic inputs and $D=71$ possible delays. We drew random independent instances of $B$ with a length of $T=1000$ time steps and with on average $1.0$ occurrences in each draw. This allowed us to generate raster plots which we use to infer $\hat{B}$. We compute the accuracy as the rate of true positive detections (both for inferring the address and its exact timing) and observe on average $\approx 98\%$ correct detections. We further extended this result by showing how the accuracy would evolve as a function of the number of simultaneous PGs, while keeping the same frequency of occurrence. We show in Figure~\ref{fig:model_results}~(a) that the accuracy of finding the right PG is still above $80\%$ accuracy with more than $1364$ overlapping PGs. Moreover, we show in Figure~\ref{fig:model_results}~(b) that (with $M=144$ PGs fixed) the accuracy increases notably as the temporal depth $D$ of the PG kernel increased, demonstrating quantitatively the computational advantage of using heterogeneous delays. These results were obtained while assuming that we know $W$. However, this is in general not the case, for instance when observing the raster plot of a population of neurons. In the following, we will define a generic visual task and determine a learning algorithm to solve that challenging task. %Inspired by the k-means algorithm, it is possible to devise a self-supervised learning algorithm. Our preliminary results show that it is possible to retrieve PGs embedded in the data, yet that further analysis is necessary to improve the convergence of the algorithm. In particular, it seems promising to use a sparseness constraint in the inference mechanism such as to remove spurious correlations in the inference.
\subsection{Extension as 3D convolutions for motion detection}
Now that our motion detection task is defined in visual space, we can extend the convolution-based detection model to a 3D convolution so that the resulting model would also benefit from spatial invariance. The use of spatio-temporal filters on an event stream has improved the performance of CNNs for an action recognition task in a model by~\citet{ghosh_spatiotemporal_2019} and here we will use a similar strategy, but focus on using an event-based representation.
For this purpose, we design 3D kernels of shape $(\Kx, \Ky, \Ktime) = (9, 9, 11)$, representing the two spatial dimensions and the range of time delays, respectively.
The computations are performed on the spatio-temporal windows that are defined by the kernels and are ``moving'' around the events, i.e. the spatial center of the spatio-temporal window around the current event $\epsilon_\arank$. Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
\begin{figure}%[h!]%[ht!]
    \includegraphics[width=\linewidth]{figures/HD-SNN.png}
    \textbf{- HD-SNN illustration. From left to right:} A 3D representation of the event stream as output of the DVS. A 3D convolution is applied, on the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. There are as many kernels as the possible number of motion classes in the recording. For clarity, only $4$ kernels are illustrated and some weights are not represented in the voxel grid. For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the non-linearity of the MLR model. The output of the MLR gives a probability for each class associated to a specific kernel to emit an output event, or spike. With an appropriate spiking mechanism, here the winner-takes-all mechanism, we obtained, as output of the HD-SNN model, a new spike train with the different spikes associated to a specific motion class. }
